 

### 2021-07-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2). However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy, perhaps due to higher RH [@zweifelMiddayStomatalClosure2002]. Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $gs$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018]. In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in the upper canopy [@carterExperimentalWarmingTropical 2021]. Higher maximum temperatures, and $g_s$ are known influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.

@@

$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. Along a vertical gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2020]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. Based on this evidence, we would expect leaves exposed to greater heat stress within canopy to have higher heat tolerance, but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.

>>>>

Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2). However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy resulting in higher $T_{leaf}$, perhaps due to higher RH [Fig. `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002]. Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018]. In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical 2021]. Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.

@@

$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorophyll a fluorescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. Along a vertical gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2020]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. Based on this evidence, we would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.
